---
title: "BiomeBGC_dataPrep"
author: 
  - Dominique Caron
  - Céline Boisvenue
date: "February 2026"
output:
  html_document:
    keep_md: yes
bibliography: citation.bib
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, eval = FALSE, results = "hold", message = FALSE)
```

# Overview

Biome-BGC is an ecophysiological model that simulates carbon, nitrogen, and water cycling in
forest ecosystems. Before it can be run, Biome-BGC needs a set of site-level inputs (e.g.,
meteorology, soil properties, species parameters) assembled into the specific `.ini` control
files that the model reads. `BiomeBGC_dataPrep` retrieves and prepares those inputs for one or
more sites (or polygons) and writes the `.ini` files needed to run Biome-BGC version 4.2 through
the module `BiomeBGC_core`.

Biome-BGC simulations happen in two stages, and this module prepares inputs for both:

- A **spin-up** run, which repeats historical climate conditions over many years until the
  simulated carbon and nitrogen pools reach a steady state. This gives the main simulation a
  realistic starting point instead of an arbitrary one.
- The **main** simulation run, which uses the spin-up's final state as its starting point and
  runs over the user's simulation period.

The user only needs to supply a study area (as one or more polygons, or points representing
site locations); all other inputs have Canada-wide defaults that the module will download and
process automatically if not supplied. Users can override any default by supplying their own
data for that input (see [Input data](#input-data) below).

# Set up

A minimal simulation needs a `studyArea`, plus a `rasterToMatch` if `studyArea` is a polygon
(points do not require one, since each point is treated as its own site). All other inputs fall
back to defaults described below.

```{r setupSim, eval = FALSE}
library(SpaDES.core)

mySim <- simInit(
  times = list(start = 2011, end = 2011),
  params = list(
    BiomeBGC_dataPrep = list(
      co2scenario = "RCP45",
      metSpinupYears = 40
    )
  ),
  modules = "BiomeBGC_dataPrep",
  objects = list(
    studyArea = studyArea, # a SpatVector of polygons or points
    rasterToMatch = rasterToMatch # required only if studyArea is a polygon
  ),
  paths = list(
    modulePath = "..",
    inputPath = "inputs",
    outputPath = "outputs"
  )
)

mySimOut <- spades(mySim)
```

# Parameters

The module's parameters fall into a few groups:

- **Spin-up controls** — `maxSpinupYears` (an upper limit on the number of simulated spin-up
  years) and `metSpinupYears` (how many years of historical meteorological data are recycled
  during spin-up).
- **Climate-change scenario** — `climModel` (which climate model to use: `"RCM4"`, `"GCM4"`, or
  `"Hadley"`), `co2scenario` (`"RCP45"` or `"RCP85"`, i.e. a lower- or higher-emissions future
  pathway), and `climateChangeOptions` (manual offsets/multipliers applied to future
  temperature, precipitation, humidity, and solar radiation, for sensitivity analyses).
- **Initial model state** — `carbonState`, `nitrogenState`, and `waterState` set the starting
  carbon, nitrogen, and water pools for the spin-up; `siteConstants` holds fixed site
  characteristics (soil depth, soil texture, elevation, latitude, albedo, N deposition, N
  fixation) that are otherwise filled in from the input data described below;
  `NDepositionLevel` controls whether atmospheric nitrogen deposition is held constant or
  allowed to vary with the CO2 scenario.
- **Output control** — `outputVariables` (which daily Biome-BGC output variables to request)
  and `savePixelGroupMap` (whether to keep the raster of pixel groups as a saved output).
- **Standard SpaDES parameters** — the `.plots`, `.plotInitialTime`, `.plotInterval`,
  `.saveInitialTime`, `.saveInterval`, `.studyAreaName`, `.seed`, and `.useCache` parameters
  follow the usual SpaDES conventions for controlling plotting, saving, and caching, shared
  across all SpaDES modules.

```{r moduleParams, echo = FALSE, eval = TRUE}
df_params <- SpaDES.core::moduleParams("BiomeBGC_dataPrep", "..")
knitr::kable(df_params)
```

# Events

The module has two event types.

## Init

The `init` event does the main work of the module, in three steps:

1. **`preparePixelGroups`** groups pixels of the study area into "pixel groups" — sets of
   pixels that share the same climate polygon, dominant species, soil properties, and
   elevation, and so can be simulated with a single Biome-BGC run. This keeps the number of
   simulations manageable for large or spatially heterogeneous study areas. The grouping is
   stored in `pixelGroupMap` and summarized in `pixelGroupParameters`.
2. **`prepareSpinupIni`** writes one spin-up `.ini` file per pixel group (and per
   scenario, where relevant), including the recycled historical meteorology, initial carbon/
   nitrogen/water state, and the requested length of the spin-up.
3. **`prepareIni`** writes the corresponding main-run `.ini` files, carrying over the
   spin-up's restart state and substituting in the main simulation's meteorological data,
   simulation period, and CO2 concentration.

## Plotting

If plotting is enabled (via the standard `.plots`/`.plotInitialTime`/`.plotInterval`
parameters), a `plot` event produces two figures, saved as PNG files under
`outputPath(sim)/BiomeBGC_figures/`:

- `climatePlot` — a time series of annual temperature and precipitation for each climate
  polygon in the study area.
- `climatePolygonMap` — a map of the climate polygons, colour-coded by polygon ID, so the user
  can see how the climate data were spatially partitioned.

## Saving

No additional objects are saved by default beyond what is controlled by the standard SpaDES
`.saveInitialTime`/`.saveInterval` parameters. If `savePixelGroupMap = TRUE`, the
`pixelGroupMap` raster is kept as a saved output rather than being discarded after use.

# Data dependencies

## Input data

Input data are the parameters driving Biome-BGC. The default input data are from various
sources, scales, and precision levels. We calibrated default ecophysiological constants for
Canadian tree species by aggregating data from @white_parameterization_2000,
@hessl_ecophysiological_2004, and the TRY plant trait database [@kattge_try_2020].
Meteorological data are retrieved through the BioSIM package [@fortin_web_2022,
@fortin_package_2022]. Site-level parameters are extracted from the following sources:

- Dominant species: NTEMS dominant species layers for the 1st year of the simulation
  [@hermosilla_characterizing_2024].
- Elevation: AWS Open Data Terrain Tiles extracted with the `elevatr` package
  [@hollister_elevatr_2025].
- N deposition: Extracted from the ADAGIO project [@robichaud_data_2025, @robichaud_data_2026].
- N fixation rate: Extracted from @reis_ely_global_2025.
- Snow pack water content: Canada monthly snow water equivalent (SWE) for January
  [@mudryk_characterization_2015].
- Soil depth: The maximum soil depth layer of NACP MsTMIP: Unified North American Soil Map
  [@liu_north_2014].
- Soil texture (sand/silt/clay percentages): Soil Landscape Grids of Canada
  [@geng_100_2025].

The only input the user must supply is `studyArea`, one or more polygons or points defining
the area(s) or site(s) to simulate. All other inputs below have Canada-wide defaults that are
downloaded and processed automatically if not supplied:

- `climatePolygons` — polygons of climate zones in which Biome-BGC can be assumed to see
  homogeneous weather (default: Canadian ecodistricts).
- `CO2concentration` — a table of atmospheric CO2 concentration (ppm) by year, consistent with
  the chosen `co2scenario`.
- `ecophysiologicalConstants` — species-specific ecophysiological constants required by
  Biome-BGC.
- `rasterToMatch` — a template raster defining the extent, resolution, and projection to use;
  required if `studyArea` is a polygon (rather than points).
- `soilTextures` — the sand/silt/clay percentage layers (see above).
- `sppEquiv` — a species-equivalency table linking the dominant species map to the
  ecophysiological constants table.

```{r moduleInputs, echo = FALSE, eval = TRUE}
df_inputs <- SpaDES.core::moduleInputs("BiomeBGC_dataPrep", "..")
knitr::kable(df_inputs)
```

## Output data

The module's main outputs are the Biome-BGC `.ini` control files, grouped by pixel group:

- `bbgcSpinup.ini` — a named list of parsed spin-up `.ini` objects (as returned by
  `BiomeBGCR::iniRead()`), one per pixel group/scenario, named by pixel group ID.
- `bbgc.ini` — a named list of parsed main-run `.ini` objects (same structure), which carry
  over the spin-up's ending state.

These parsed `.ini` objects are passed on to `BiomeBGC_core`, which writes them to disk (via
`BiomeBGCR::iniWrite()`) and runs Biome-BGC itself. The module also returns the spatial units
used to generate those files:

- `pixelGroupMap` — a raster showing which pixel group each pixel belongs to (only kept as a
  saved output if `savePixelGroupMap = TRUE`).
- `pixelGroupParameters` — a table of the site characteristics (climate polygon, dominant
  species, soil, elevation, etc.) associated with each pixel group.

```{r moduleOutputs, echo = FALSE, eval = TRUE}
df_outputs <- SpaDES.core::moduleOutputs("BiomeBGC_dataPrep", "..")
knitr::kable(df_outputs)
```

# Links to other modules

`BiomeBGC_dataPrep` is typically the first module run in a Biome-BGC workflow — it has no
upstream module dependencies, and prepares inputs for `BiomeBGC_core`, which reads the
`bbgcSpinup.ini`/`bbgc.ini` files produced here to run the model. Model output from
`BiomeBGC_core` can in turn be validated against observed flux tower data using
`BiomeBGC_validationFluxTower`.

```mermaid
flowchart LR
    A[studyArea + optional inputs] --> B[preparePixelGroups]
    B --> C[prepareSpinupIni]
    C --> D[prepareIni]
    D --> E["bbgcSpinup.ini / bbgc.ini"]
    E --> F[BiomeBGC_core]
    F --> G[BiomeBGC_validationFluxTower]
```

# References
